Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-29 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Merge old and new methods sections

\keywords{Iterative imputation; non-convergence; mice}

@@

Currently, the recommended practice for evaluating convergence in
iterative imputation algorithms is to visually inspect imputations for
signs of non-convergence. \textbf{(Add chain means and variances here
already? Because it's insufficient that these are univariate?)}. This
method is insufficient on two counts: 1) it may be challenging to the
untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred. Yet, monitoring convergence of iterative imputation
algorithms diagnostically is challenging. Iterative imputation
algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
MCMC methods, convergence is not from a scalar to a point but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Therefore, the aim of
convergence diagnostics for MCMC methods is not to establish the point
at which convergence is reached, but to monitor signs of
non-convergence. Several of such diagnostics exist for MCMC methods, but
it is not known whether these are appropriate for iterative imputation
algorithms.

@@

\textbf{Include sub-questions?:} How can non-convergence be identified
diagnostically? Are common MCMC non-convergence diagnostics appropriate
for MICE? And if so, which threshold should be used to diagnose
non-convergence? How many iterations are sufficient/needed to be able to
diagnose non-convergence? Are the default number of iterations
sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
\textbf{How severe is it when the algorithm has not converged? And what
are guidelines for practice? Can the parameter of interest, estimand
\(Q\), be correct when the algorithm is not (yet) converged, and vice
versa?}

@@

There are two requirements for convergence of iterative algorithms:
mixing and stationarity \citep{gelm13}. Without mixing, imputation
chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
Without stationarity, there is trending within imputation chains, which
implies that further iterations would yiled a different set of
imputations. In Figure \ref{fig:non-conv}\ldots{} \textbf{Explain what
we see, namely example by \citet{buur18} reproduced, showing the
traceplots of chain means for some variable. The first plot is typical
convergence of MICE, the second is pathological non-convergence because
of a mis-specified imputation model. Each line is an imputation. In the
first plot, the chains intermingle nicely and there is little to no
trending. In the second plot, there is a lot of trending and some chains
do not intermingle. Importantly, the chain means at the last iteration
(the imputed value per \(m\)) are very different between the two plots.
The algorithm with the mis-specified model yields imputed values that
are on average a magnitude two larger than those of the typically
converged algorithm. This shows the importance of reaching converged
states in iterative imputation algorithms.}

@@

\hypertarget{potential-scale-reduction-factor}{%
\subsection{Potential scale reduction
factor}\label{potential-scale-reduction-factor}}

An updated version of \(\widehat{R}\) has been proposed by
\citet{veht19}, (p.~5). \textbf{This version may be suitable for
iterative imputation.} Let \(M\) be the total number of chains, \(T\)
the number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

\hypertarget{autocorrelation}{%
\subsection{Autocorrelation}\label{autocorrelation}}

@@

\hypertarget{in-practiceanswer}{%
\subsection{In practice/Answer}\label{in-practiceanswer}}

\textbf{To assess whether \(\widehat{R}\) and \(AC\) may be appropriate
non-convergence identifiers for iterative imputation algorithms, the
following condition must hold when applied on the two algorithms plotted
above: the methods indicate worse performance for the mis-specified
model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
and \(AC\)-values than the typical performance). And additionally the
methods should reflect the increasing convergence in the typical
convergence situation as the number of iterations goes up.}

\textbf{In figure \ref{fig:diagnostics}A (}add panel labels\textbf{),
the chain means are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

\caption{Convergence diagnostics pathological non-convergence versus typical convergence. **ADD PANEL LABELS**}\label{fig:diagnostics}
\end{figure}

\hypertarget{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}{%
\subsection{Simulation Hypothesis (in bullets!! maybe move to simulation
section?)}\label{simulation-hypothesis-in-bullets-maybe-move-to-simulation-section}}

For multiple imputation algorithms, it holds that convergence is reached
when there is no dependency between subsequent iterations of imputation
chains (\(AC = 0\)), and chains intermingle such that the only
difference between the chains is caused by the randomness induced by the
algorithm (\(\widehat{R} = 1\)). \textbf{We expect that a completely
converged state will not be reached.}

This study evaluates whether \(\widehat{R}\) and \(AC\) could diagnose
convergence of multiple imputation algorithms. We assess the performance
of the two convergence diagnostics against the recommended evaluation
criteria for MI methods \citep[i.e., average bias, average confidence
interval width, and empirical coverage rate across simulations;][\(\S\)
2.5.2]{buur18}. \textbf{That is, there is no baseline measure available
to evaluate performance against.}

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
after at least 50 iterations. Therefore\ldots{} Vehtari's proposed
threshold of \(\widehat{R} < 1.01\) will then be too stringent for
diagnosing convergence. This over-estimation may, however, be diminished
because \(\widehat{R}\) can falsely diagnose convergence if initial
values of the algorithm are not appropriately over-dispersed
\citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial values are chosen
randomly from the observed data. Therefore, we cannot be certain that
the initial values are over-dispersed. We expect this to have little
effect on the hypothesized performance of \(\widehat{R}\). \textbf{(Add
actual hypothesis: we thought that mice would converge sooner than
\(\widehat{R}\) would indicate that it did)} We hypothesize that high AC
values are implausible in converged MI algorithms. \textbf{Added later:
High \(AC\)-values are implausible in MI procedures. That is, the
randomness induced by the MI algorithm effectively mitigates the risk of
dependency within chains. But this doesn't seem the case in practice??}

@@

\hypertarget{how-to-diagnose-non-convergence}{%
\subsection{How to diagnose
non-convergence?}\label{how-to-diagnose-non-convergence}}

\textbf{Move to section above} Mixing and stationarity have histroically
been inspected visually, by evaluating traceplots of scalar summaries of
interest (\(\theta\)s; e.g., chain means and chain variances). As
\citet{buur18} describes, users can also specify a model-specific scalar
summary (e.g., a regression coefficient). A user-specified scalar
summary, however, is not universal to all complete data problems.
Therefore, as inspired by \citep{mack03}, we propose a \(\theta\) that
summarizes the multivariate state space of the algorithm. Namely, the
first eigenvalue of the variance-covariance matrix of the \(M\)
completed datasets. \textbf{The first eigenvalue has the appealing
property that is not dependent on the substantive model of interest.}

The methods under evaluation are \(\widehat{R}\) and \(AC\). These will
be computed for the \textbf{four} scalar summaries of interest: chain
means (i.e., per imputation, per variable: on \(y_{imp, m, j}\)), chain
variances (i.e., per imputation, per variable: on \(y_{imp, m, j}\)),
and the first eigenvalue of the variance-covariance matrix across
imputations (i.e., per imputation: on \{\(y_{obs}, y_{imp, m}\)\})
\textbf{add user-specified \(\theta\): estimated regression coefficient
per imputation}.

The performance of two methods (\(\widehat{R}\) and \(AC\)) on each of
these scalar summaries of interest will be evaluated with several
estimands, i.e.~quantities of scientific interest \(Q\). It is assumed
that when \(\bar{Q}\) (the pooled result across imputations) is an
unbiased and confidence valid estimate of \(Q\), the algorithm is
sufficiently converged. For each scalar \(\theta\), one estimand \(Q\)
is defined that may be of interest in empirical research. The
\(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s are
evaluated against the bias in univariate mean estimates (\textbf{all
variables or just for Y??}). Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

\textbf{Move this to Results:} As expected, conditions with a higher
proportion of missingness and/or a lower number of iterations show more
signs of non-convergence, as indicated by more extreme bias in the
estimated \(Q\)s. Roughly speaking this means that MICE is indeed not
converged as \(t=1\), and converges gradually as \(t\) increases. The
point at which an additional iteration does not lead to an improvement
of the estimates depends on the difficulty of the missingness
problem---between XYZ and XYZ. This means that the algorithm has
converged sufficiently (\textbf{under the current specifications}).

As required for a method to diagnose non-convergence, the values of
\(\widehat{R}\) follow a trend similar to the performance measures.
\(\widehat{R}\) values are generally lower in conditions with a higher
number of iterations, and somewhat higher in conditions with a higher
percentage of missing data. \textbf{Decide where to specify the two
calculations of AC:} Initially, autocorrelation did not show a decrease
with an increasing number of iterations. But, after replacing the ACF
function with a manual calculation for autocorrelation, the
autocorrelation values were indeed decreasing with a higher number of
iterations.

\hypertarget{when-to-diagnose-non-convergence}{%
\subsection{When to diagnose
non-convergence?}\label{when-to-diagnose-non-convergence}}

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely
thresholds for MCMC algorithms, because of its convergence to a
distribution. In practice, non-convergence is usually diagnosed when
\(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is
performed to assess whether \(AC\) is significantly different from zero.
Performance measures are the same as above: unbiased, confidence valid
estimates.

\textbf{Move this to Results:} As expected, complete convergence
(\(\widehat{R}=1\) and \(AC=0\)) is never observed. But what about the
thresholds for practice? \(\widehat{R}\) \textless1.2 is not stringent
enough, because conditions in which \(\widehat{R}\) is smaller than 1.2
do not all yield unbiased estimates of the \(Q\)s. Moreover, there is a
dip in the \(\widehat{R}\) values, after which \(\widehat{R}\)
\textgreater{} 1.2 again. If the algorithms would be terminated at the
iteration where \(\widehat{R}\) \textless{} 1.2 occurs for the first
time, the increase in \(\widehat{R}\) values might be missed. The
threshold \(\widehat{R}\) \textless{} 1.1 is somewhat conservative in
comparison with the performance measures. But it seems OK.
\(\widehat{R}\) \textless{} 1.01 is too stringent compared to the
performance measures. And it is not obtained for any number of
iterations (\textbf{but may be necessary in more complex complete data
models???}).

\hypertarget{start-old-methods-section}{%
\subsection{Start old methods section}\label{start-old-methods-section}}

\textbf{Add to this section: 1) explicit mention of simulation
conditions; 2) emphasize that the plots are averages across repetitions,
not within MICE; 3) sample effects due to single complete dataset}
Convergence of the MICE algorithm is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence by: 1) terminating the MICE algorithm at different
imputation chain lengths (\(t = 1, 2,..., 100\)), and 2) varying the
missingness proportions (\(p_{mis} =.05,.25,.50,.75,.95\)). The
assumption underlying the different number of iterations is that the
algorithm generally does reach convergence at \(t=1\), because the
initial values are sampled randomly from the set of observed datapoints.
\textbf{As the number of iterations goes up, the imputation chains will
become independent of the initial values until the point at which an
extra added iteration does not lead to a more converged state.} The
second set of experimental conditions under consideration--the
missingness proportions--are chosen to reflect the difficulty of the
missingness problem. The inherent assumption is that low missingness
proportions will lead to quick algorithmic convergence, since there is a
lot of information in the observed data. Higher missingness proportions
then cause slower convergence. However, if the fraction of missing
information is very high, there is so little information in the data
that the random component in the algorithm will take the overhand and a
stable but very uncertain (high variance) point will be reached.

The data-generating mechanism is a multivariate normal distribution,
representing person-data on three predictor variables (from an
unspecified social scientific field of study). Let our predictor space
be defined by the following a multivariate normal distribution

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} In this study, sampling variance is not of interest.
Therefore, a single complete set may serve as comparative truth in all
simulation runs \citep{vink14}. A finite population of \(N=1000\) is
simulated using the \texttt{mvtnorm} package \citep{mvtnorm}.
Subsequently, a fourth variable is constructed to serve as outcome
variable in a multiple linear regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

\textbf{We consider four quantities of scientific interest
\citep[`conceptual estimands';][]{morr19}, namely the descriptive
statistics of all variables (mean and standard deviation), \(\beta_2\),
and the percentage of variance in \(Y\) explained (coefficient of
determination; \(R^2\)). We solve a multiple linear regression problem,
where outcome \(Y\) is regressed on predictors \(X_1\), \(X_2\) and
\(X_3\)}

@@

The complete data is \emph{amputed} once for each simulation repetition
with function \texttt{mice::ampute()}.

(\textbf{change this to reflect current simulation set-up with 5, 25,
50, 75, and 95\% of cases having missing data: The missingness is
univariate, and the probability to be missing is the same for all four
variables, namely 20\% (\texttt{prop\ =\ 0.8,\ mech\ =\ "MCAR"}). This
leaves 20\% of the rows completely observed). }

\textbf{Maybe move to discussion:}Proper performance of the convergence
diagnostics under at least missing completely at random (MCAR)
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
may not be extrapolated to other missingness mechanisms.
\textbf{Convergence diagnostics should therefore at least apply to the
MCAR situation, before more complex missingness should be explored.}

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)). Each
repetition of the simulation thus starts from 1 complete dataset, then
splits into 5 missingness conditions (\(p_{mis} =.05,.25,.5,.75,.95\)),
each of the amputed datasets is imputed 5 times (\(m = 1, 2,..., 5\))
for each iteration condition (\(t = 1, 2,..., 100\)). From these
\(5\times5\times100\) imputations, we extract the \(\theta\)s to apply
\(\widehat{R}\) and \(AC\) on (\textbf{i.e., chain means, chain
variances, \(\beta\)s per imputation, and the first eigenvalue of the
variance-covariance matrix per imputation}). We calculate
\(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19}
recommendations, and \(AC\) as the correlation between the \(t^{th}\)
and the \((t+1)^{th}\) iteration. For each \(\theta\) we estimate the
corresponding scientific estimand \(Q\).

@@

\hypertarget{old-results-section}{%
\section{Old Results section}\label{old-results-section}}

@@

>>>>

\keywords{missing data, iterative imputation, non-convergence, mice}

@@

Currently, the recommended practice for evaluating convergence in
iterative imputation algorithms is to visually inspect imputations for
signs of non-convergence. \textbf{Add that this refers to chain means
and variances already? It's insufficient that these are univariate, so
multivariate state space of the algorithm may not be converged when
univariates are ok. And \citet{buur18}`s suggestion to track
user-defined statistics (e.g., regression coefficient) may be somewhat
advanced for empirical researchers. Moreover, this is not
model-independent, i.e., only apply to one substantive model, while one
of the advantages of MI is that missing data problem and scientific
problem are split. Then we can say 'we propose a novel\ldots{}' and
integrate this: Note that usually we only evaluate the convergence of
univariate scalar summaries (e.g., chain means or variances). With these
we cannot diagnose convergence of multivariable statistics (i.e.,
relations between scalar summaries). Van Buuren \citeyearpar[\(\S\)
4.5.2]{buur18} proposed to implement multivariable evaluation of the
MICE algorithm through eigenvalue decomposition building on the work of
\citet{mack03}. Eigenvalues of a covariance matrix are a measure of the
data's covariance.}.

This method is insufficient on two counts: 1) it may be challenging to
the untrained eye, and 2) only severely pathological cases of
non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18}.
Therefore, a quantitative, diagnostic evaluation of convergence would be
preferred. Yet, monitoring convergence of iterative imputation
algorithms diagnostically is challenging. Iterative imputation
algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In
MCMC methods, convergence is not from a scalar to a point but from one
distribution to another. The values generated by the algorithm (e.g.,
imputed values) will vary even after convergence. Therefore, the aim of
convergence diagnostics for MCMC methods is not to establish the point
at which convergence is reached, but to monitor signs of non-convergence
\citep{hoff09}. Several of such diagnostics exist for MCMC methods, but
it is not known whether these are appropriate for iterative imputation
algorithms.

@@

\textbf{Include sub-questions!} How can non-convergence be identified
diagnostically? Are common MCMC non-convergence diagnostics appropriate
for MICE? And if so, which threshold should be used to diagnose
non-convergence? How many iterations are sufficient/needed to be able to
diagnose non-convergence? Are the default number of iterations
sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)?
\textbf{How severe is it when the algorithm has not converged? And what
are guidelines for practice? Can the parameter of interest, estimand
\(Q\), be correct when the algorithm is not (yet) converged, and vice
versa?} \emph{(Maybe add these too? What are the effects of continued
iterations on estimates, predictions and inferences? Do the answers
differ with varying missingness proportions? That is, we vary the nr of
iterations and the missingness proportion because we assume that the
algorithm has not reached convergence at \(t=1\), and performs worse
with moderate to high missingness (not so much with little or a LOT of
missingness).)}

@@

Mixing and stationarity have histroically been inspected visually, by
evaluating traceplots of scalar summaries of interest (\(\theta\)s;
e.g., chain means and chain variances). As \citet{buur18} describes,
users can also specify a model-specific scalar summary (e.g., a
regression coefficient). A user-specified scalar summary, however, is
not universal to all complete data problems. Therefore, as inspired by
\citep{mack03}, we propose a \(\theta\) that summarizes the multivariate
state space of the algorithm. Namely, the first eigenvalue of the
variance-covariance matrix of the \(M\) completed datasets. \textbf{The
first eigenvalue has the appealing property that is not dependent on the
substantive model of interest.}

\textbf{Since convergence of iterative imputation algorithms is in
distribution, there is not a unique point at which the algorithm reaches
a converged state. We can only evaluate scalar summaries of the
multivariate state space of the algorithm, \(\theta\)s. The recommended
scalar summaries to evaluate are chain means and chain variances.
Additionally, researchers may ``monitor some statistic of scientific
interest'' \citep[\(\S\) 6.5.2]{buur18}. Because of the issues mentioned
above, we propose a novel \(\theta\) to monitor: the first eigenvalue of
the variance-covariance matrix of the completed data. Let
\(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues
of \(\Sigma\) in each of the \(M\) completed datasets
\({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarises the
covariances in the completed datasets.}

There are two requirements for convergence of iterative algorithms:
mixing and stationarity \citep{gelm13}. Without mixing, imputation
chains do not intermingle nicely, \textbf{indicating that \ldots{}}.
Without stationarity, there is trending within imputation chains, which
implies that further iterations would yiled a different set of
imputations.

To illustrate what non-convergence looks like in MI, we reproduce the
example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure
\ref{fig:non-conv} shows the traceplot for one of the variables in the
example. we see the average of the imputed values for a variable \(j\)
in \(y_{imp}\). The average imputed values in the left hand plot are
about a magnitude 2 lower than the right hand plot. So non-convergence
has a impact on the average imputed value for \(j\) in \(y_{imp, m}\).
This difference (presumably) translates to bias in the pooled estimate
\(\bar{\mu}_j\) as well. And maybe also to higher order statistics,
e.g.~variances, covariances, etc. Therefore, it's important to converge.
\textbf{Explain what we see, namely example by \citet{buur18}
reproduced, showing the traceplots of chain means for some variable. The
first plot is typical convergence of MICE, the second is pathological
non-convergence because of a mis-specified imputation model. Each line
is an imputation. In the first plot, the chains intermingle nicely and
there is little to no trending. In the second plot, there is a lot of
trending and some chains do not intermingle. Importantly, the chain
means at the last iteration (the imputed value per \(m\)) are very
different between the two plots. The algorithm with the mis-specified
model yields imputed values that are on average a magnitude two larger
than those of the typically converged algorithm. This shows the
importance of reaching converged states in iterative imputation
algorithms.}

\hypertarget{methods-under-evaluation}{%
\subsection{Methods under evaluation}\label{methods-under-evaluation}}

@@

\hypertarget{potential-scale-reduction-factor}{%
\subsubsection{Potential scale reduction
factor}\label{potential-scale-reduction-factor}}

An updated version of \(\widehat{R}\) has been proposed by
\citet{veht19} (p.~5). \textbf{This version may be suitable for
iterative imputation.} Let \(M\) be the total number of chains, \(T\)
the number of iterations per chain, and \(\theta\) the scalar summary of
interest (e.g., chain mean or chain variance). For each chain
(\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and
average these to obtain within-chain variance \(W\).

@@

\hypertarget{autocorrelation}{%
\subsubsection{Autocorrelation}\label{autocorrelation}}

@@

\hypertarget{in-practice}{%
\subsection{In practice}\label{in-practice}}

\textbf{Before using them in the simulation, the two methods show at
least be able to distinguish between the two algorithms plotted in
Figure 2. From visual inspection we know that the typical convergence
has some signs of non-mixing around iteration 2, and little trending. In
the non-convergence situation, there is a lot of trending upto iteration
6, after which the chains reach a somewhat more stationary state. Mixing
in this situation gets worse from the first iteration onward, and
gradually gets better around iteration 7. To assess whether
\(\widehat{R}\) and \(AC\) may be appropriate non-convergence
identifiers for iterative imputation algorithms, the following condition
must hold: the methods indicate worse performance for the mis-specified
model with pathological non-convergence (i.e., higher \(\widehat{R}\)-
and \(AC\)-values than the typical performance). And additionally the
methods should reflect the increasing convergence in the typical
convergence situation as the number of iterations goes up.}

\textbf{In Figure \ref{fig:diagnostics}A, the chain means from Figure
\ref{fig:non-conv} are plotted again, now together---as are the
non-convergence diagnostics. Panel B shows \(\widehat{R}\) as computed
by implementing \citet{veht19} 's recommendations. As required,
\(\widehat{R}\) indicates less signs of non-convergence as the number of
iterations goes up }in the typical convergence situation\textbf{. The
superior performance of the typical convergence over the pathological
non-convergence is less prominent, even flipped for \(t<4\). Panel C
displays the \(AC\) as computed with the R function
\texttt{stats::acf()}. When we look at this panel, we conclude something
weird. The \(AC\)-values indicate equal performance (up-to \(t=5\)) for
the typical convergence and the pathological non-convergence, while
there is obvious trending in the latter. Moreover, the best convergence
(as indicated by the lowest \(AC\)-value) is observed at \(t=2\), but
looking at the chain means in panel A, there should be some signs of
trending up-to iteration number seven. After consulting the
documentation on stats::acf() \citep{R}, we conclude that this
implementation of \(AC\) is not suitable for iterative imputation
algorithms. There is a correction factor for a mathematical shortcut
that works in the limit. According to \citet{box15}, the function works
when the number of iterations \textgreater{} 50. The default number of
iterations in iterative imputation, however, is often much lower.
Therefore, we compute \(AC\) manually, see panel D. The AC values in
this plot do meet the requirements (\ldots) and will therefore be used
in the simulation study.}

@@

\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
\end{figure}

@@

\textbf{Add to this section: 1) explicit mention of simulation
conditions; 2) emphasize that the plots are averages across repetitions,
not within MICE; 3) sample effects due to single complete dataset}

The aim of the simulation study is to evaluate the impact of inducing
non-convergence on several quantities of scientific interest \(Q\), and
to subsequently evaluate the performance of \(\widehat{R}\) and \(AC\)
in identifying the non-convergence.

Non-convergence will be induced by: 1) varying the missingness
proportions in \(y\) (\(p_{mis} =.05,.25,.50,.75,.95\)), and 2)
terminating the MICE algorithm at different imputation chain lengths
(\(t = 1, 2,..., 100\)). The assumption underlying the different number
of iterations is that the algorithm generally does reach convergence at
\(t=1\). The initial values in the first iteration are sampled randomly
from the set of observed datapoints. \textbf{As the number of iterations
goes up, the imputation chains become independent of the initial values
until the point at which an extra added iteration does not lead to a
more converged state.} The second set of experimental conditions under
consideration--the missingness proportions--are chosen to reflect the
difficulty of the missingness problem. The inherent assumption is that
low missingness proportions will lead to quick algorithmic convergence,
since there is a lot of information in the observed data. Higher
missingness proportions then cause slower convergence. However, if the
fraction of missing information is very high, there is so little
information in the data that the random component in the algorithm will
take the overhand and a stable but very uncertain (high variance) state
will be reached.

\hypertarget{hypothesis}{%
\subsection{Hypothesis}\label{hypothesis}}

We hypothesize that simulation conditions with a higer \(p_miss\) and
lower \(t\) have inferior convergence. We assume that when \(\bar{Q}\)
is an unbiased and confidence valid estimate of \(Q\), the algorithm is
\emph{sufficiently} converged.

\textbf{(bruggetje hier)} For multiple imputation algorithms, it holds
that convergence is reached when there is no dependency between
subsequent iterations of imputation chains (\(AC = 0\)), and chains
intermingle such that the only difference between the chains is caused
by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
\textbf{We expect that a completely converged state will not be
reached.}

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
after at least 50 iterations. Therefore we hypothesize that Vehtari's
proposed threshold of \(\widehat{R} < 1.01\) will be too stringent for
diagnosing convergence in iterative imputation. This over-estimation
may, however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\).
\textbf{(Add actual hypothesis: we thought that mice would converge
sooner than \(\widehat{R}\) would indicate that it did)} We hypothesize
that high AC values are implausible in converged MI algorithms.
\textbf{Added later: High \(AC\)-values are implausible in MI
procedures. That is, the randomness induced by the MI algorithm
effectively mitigates the risk of dependency within chains. But this
doesn't seem the case in practice??}

\hypertarget{set-up}{%
\subsection{Set-up}\label{set-up}}

Convergence of the MICE algorithm is investigated through model-based
simulation in \texttt{R} \citep[version 3.6.3;][]{R}. The simulation
set-up is summarized in the pseudo-code below. The complete \texttt{R}
script of the simulation study is available from
\href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}.

@@

The data-generating mechanism is a multivariate normal distribution,
representing person-data on three predictor variables (from an
unspecified social scientific field of study). Let our predictor space
be defined as \begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*}

In this study, sampling variance is not of interest. Therefore, a single
complete set may serve as comparative truth in all simulation runs
\citep{vink14}. A finite population of \(N=1000\) is simulated using the
\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
variable is constructed as outcome variable in a multiple linear
regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

The complete data is \emph{amputed} once for each simulation repetition
with function \texttt{mice::ampute()}. The missingness is multivariate,
and the probability to be missing is the same for all \(n \times k\)
cells in \(y\) \citep[`MCAR';][]{rubin87}. Proper performance of the
convergence diagnostics under at least `missing completely at random'
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
may not be extrapolated to other missingness mechanisms. The missinness
proportion varies between conditions (\(p_{mis} =.05,.25,.5,.75,.95\)).

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).

The methods under evaluation, \(\widehat{R}\) and \(AC\), are computed
for four scalar summaries \(\theta\): chain means (i.e., the mean in
each \(y_{imp, m, j}\)), chain variances (i.e., the variance in each
\(y_{imp, m, j}\)), the statistic of scientific interest \(\hat{Q}\) in
each \{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
\(\lambda_1\). We calculate \(\widehat{R}\) by implementing Vehtari et
al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration. For
each \(\theta\) we estimate the corresponding scientific estimand \(Q\).

The performance of two methods on each of these scalar summaries will be
evaluated with several estimands, i.e.~quantities of scientific interest
\(Q\). For each \(\theta\), one estimand \(Q\) is defined that may be of
interest in empirical research. The \(\widehat{R}\) and \(AC\) values
with chain means as \(\theta\)s are evaluated against the bias in
univariate mean estimates. Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

We consider four quantities of scientific interest, namely the
descriptive statistics of all variables (mean and standard deviation),
\(\beta_2\), and the percentage of variance in \(Y\) explained
(coefficient of determination \(R^2 \times 100\)). We solve a multiple
linear regression problem, where outcome \(Y\) is regressed on
predictors \(X_1\), \(X_2\) and \(X_3\)

@@

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

As expected, conditions with a higher proportion of missingness and/or a
lower number of iterations show more signs of non-convergence, as
indicated by more extreme bias in the estimated \(Q\)s. Roughly speaking
this means that MICE is indeed not converged as \(t=1\), and converges
gradually as \(t\) increases. The point at which an additional iteration
does not lead to an improvement of the estimates depends on the
difficulty of the missingness problem---between XYZ and XYZ. This means
that the algorithm has converged sufficiently (\textbf{under the current
specifications}).

As required for a method to diagnose non-convergence, the values of
\(\widehat{R}\) follow a trend similar to the performance measures.
\(\widehat{R}\) values are generally lower in conditions with a higher
number of iterations, and somewhat higher in conditions with a higher
percentage of missing data. Autocorrelation values were indeed
decreasing with a higher number of iterations.
